### 2022-08-13 draft.tex
Deploying to gh-pages from @ FrancisBanville/ms_maxent_networks@647f4ae1f1a69dd22182f068828fcd142e1f4afe 🚀

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

The principle of maximum entropy is a robust mathematical method of
finding least-biased probability distributions that have some desired
properties given by prior knowledge on a system. We first used MaxEnt to
predict the joint degree distribution and the degree distribution of
maximum entropy given known numbers of species and of interactions. We
found that the resulting joint degree distributions were more symmetric
than the ones of empirical food webs, which suggests that other
constraints might be needed to improve those predictions. We also used
MaxEnt to predict networks of maximum entropy with a specified
structure. These networks are the most complex, or random, given the
constraints used. Likewise, we found that knowledge of species richness
and of the number of interactions were not sufficient to reproduce many
aspects of network structure found in nature. However, a model based on
the entire joint degree sequence, i.e.~on the number of prey and
predators for each species, gave more convincing results. Our type II
MaxEnt network model yielded better or similar predictions than the type
II null model, also based on the joint degree sequence, for most
measures considered including the motifs profile. This suggests that the
joint degree sequence drives many aspects of network structure. Indeed,
considering our findings that empirical networks are close to maximum
entropy for a given joint degree sequence, our results suggest that
food-web topology is configured almost entirely randomly around these
marginal numbers of predators and prey.

Our results bring to the forefront the role of the joint degree
distribution in shaping food-web structure. This echoes the work of
Williams (2011), which shows using MaxEnt how the degree distribution
can drive the structure of bipartite networks. Fortuna et al. (2010)
also found that the degree distribution of ecological networks is
associated with some of their emerging properties such as nestedness and
modularity. Network ecologists tend to focus on several emerging
properties of food webs (Delmas et al. 2019), although we show that
food-web structure is very well predicted by the joint degree
distribution. In the light of our results, there is a lot more
ecological information in the difference between properties of empirical
networks and MaxEnt predictions than in these properties themselves.

@@

There are different ways to generate food webs using MaxEnt. In this
contribution, we used a method based on simulated annealing to find the
network configuration with the highest SVD entropy while fixing some
aspects of its structure. Another technique, also based on simulated
annealing, could start by generating a food web randomly with fixed
numbers of species and interactions and calculating its joint degree
distribution. Pairs of interactions could be swapped sequentially until
we minimize the divergence between the calculated joint degree
distribution and the one of maximum entropy obtained analytically. In
that case, this is the entropy of the joint degree distribution that
would be maximized, not the one of the network's topology. To a certain
extent, this method would thus bridge the gap between the two approaches
presented in this article (i.e.~analytical and heuristic MaxEnt models).
More research is needed to compare the quality of different methods
generating adjacency matrices of food webs using MaxEnt.

Another method generating adjacency matrices are maximum entropy graph
models, which predict a probability distribution of networks under soft
or hard constraints (e.g., Park and Newman 2004; Cimini et al. 2019).
Under hard constraints, every network with a non-zero probability
exactly satisfies the constraints on its structure. This is in contrast
with soft constraints, which require that networks satisfy them on
average (i.e.~many networks with a non-zero probability do not have the
exact structure set by the constraints). Maximum entropy graph models
are helpful because they can provide probability distributions for many
network properties, by measuring the structure of all adjacency matrices
with a non-zero probability. However, we consider that our approach
based on simulated annealing is more flexible and more computationally
efficient. Indeed, many measures of food-web structure are hard to
translate into mathematical constraints. Moreover, because food webs are
directed networks with self-loops, it makes the mathematical derivation
of maximum entropy graph models difficult. We believe that identifying
heuristically what really constrains the topology of food webs is a
useful first step before attempting to derive the mathematical
formulation of a maximum entropy graph model adapted to food webs.

Even though we decided to work with point estimates, it is possible to
have a (non MaxEnt) probability distribution of networks when working
with the entire chains (or part of the chains).

@@

Our method and results could be used for different purposes. First, they
could be used as first-order approximations of network structure when
only state variables are known. This could prove useful when predicting
network structure at large spatial scales, where few ecological
information is known at that scale. Second, they could be used as
informative priors in Bayesian analyses of the structure of ecological
networks (e.g., Cirtwill et al. 2019). Third, they could be used to make
better predictions of pairwise species interactions by constraining the
space of feasible networks, as discussed in Strydom et al. (2021).
Finally, they could be used as alternative null models of ecological
networks to better understand ecological mechanisms driving food-web
structure. In that case, our model might need to be slightly adapted to
give a probability distribution of Boolean networks (in contrast with
point estimates of maximum entropy networks).

\hypertarget{conclusion}{%
\subsection{Conclusion}\label{conclusion}}

One of the biggest challenges in using the principle of maximum entropy
is to identify the set of state variables that best reproduce empirical
data. We found that the numbers of prey and predators for each species
are important state variables for the prediction of maximum entropy
networks. However, our predictions overestimated some measures of
network structure, especially the maximum trophic level and network
diameter. Therefore, we should continue playing the ecological detective
to find these other topological constraints that would improve the
predictions of our MaxEnt network models.

@@

>>>>

\hypertarget{conclusion}{%
\section{Conclusion}\label{conclusion}}

The principle of maximum entropy is a robust mathematical method of
finding least-biased probability distributions that have some specified
properties given by prior knowledge about a system. We first applied
this conventional MaxEnt approach on food webs to predict species level
properties, namely the joint degree distribution and the degree
distribution of maximum entropy given known numbers of species and
interactions. We found that the joint degree distributions of maximum
entropy had a similar shape to the ones of empirical food webs in
high-connectance systems. However, these MaxEnt distributions were more
symmetric than the ones of empirical food webs when connectance was low,
which suggests that other constraints might be needed to improve these
predictions in low-connectance systems. Then, we used a slightly
different approach that aimed at finding heuristically the network
configuration with the highest SVD entropy, i.e.~whose vector of
relative singular values has maximum entropy. This network of maximum
entropy is the most complex, or random, given the specified structure.
We found that the heuristic maximum entropy model based on connectance
did not predict the structure of sampled food webs very well. However,
the heuristic maximum entropy model based on the entire joint degree
sequence, i.e.~on the number of prey and predators for each species,
gave more convincing results. Indeed, this model reproduced food-web
structure surprisingly well, including the highly informative motifs
profile. Nevertheless, it was not able to predict realistic food-chain
lengths.

Our results bring to the forefront the role of the joint degree
distribution in shaping food-web structure. This echoes the work of
Fortuna et al. (2010), who found that the degree distribution of
ecological networks drive their emerging structure such as their
nestedness and modularity. Network ecologists tend to focus on several
measures of food webs when studying the ecological consequences of their
structure (McCann 2011; Delmas et al. 2019). In fact, following Williams
(2011), we believe there is a lot more ecological information in the
deviation between these properties in empirical systems and in their
maximum entropy configuration given a fixed joint degree sequence.

@@

In this contribution, we used a method based on simulated annealing to
find the network configuration with the highest SVD entropy while fixing
some aspects of its structure. However, there are different ways to
generate adjacency matrices using MaxEnt. Another technique, also based
on simulated annealing, could begin by generating a food web randomly
with fixed numbers of species and interactions and calculating its joint
degree distribution. Pairs of interactions could then be swapped
sequentially until we minimize the divergence between the calculated
joint degree distribution and the one of maximum entropy obtained
analytically. In that case, this is the entropy of the joint degree
distribution that would be maximized, not the one of the network's
topology. To a certain extent, this method would thus bridge the gap
between the analytical and heuristic approaches presented in this
article. More research is needed to compare the quality of different
methods generating adjacency matrices of food webs using MaxEnt.

Maximum entropy graph models are another type of methods that predict a
distribution of adjacency matrices under soft or hard constraints (e.g.,
Park and Newman 2004; Cimini et al. 2019). Under hard constraints, every
network with a non-zero probability exactly satisfies the constraints on
its structure. This is in contrast with soft constraints, which require
that networks satisfy them on average (i.e.~many networks with a
non-zero probability do not have the exact structure set by the
constraints). Maximum entropy graph models are helpful because they can
provide probability distributions for many network properties by
measuring the structure of all adjacency matrices with a non-zero
probability. However, we consider that our approach based on simulated
annealing is more flexible and more computationally efficient. Indeed,
many measures of food-web structure are hard to translate into
mathematical constraints. Moreover, because food webs are directed
networks that can have self-loops, it makes the mathematical derivation
of maximum entropy graph models difficult. We believe that identifying
heuristically what really constrains the topology of food webs is a
useful first step before attempting to derive the mathematical
formulation of a maximum entropy graph model for food webs.

@@

Our analytical and heuristic models can be applied for different
purposes. First, they could be used to generate first-order
approximations of a network's properties when state variables are known
empirically. For example, knowing the number of species in an ecological
community, we can predict its number of interactions using the flexible
links model and then predict its joint degree distribution with minimal
biases using the principle of maximum entropy. This could prove
particularly useful when predicting network structure at large spatial
scales, subdividing the study area into smaller communities (e.g., grid
cells). Indeed, because species richness and other ecological data are
increasingly abundant (e.g., Dickinson, Zuckerberg, and Bonter 2010),
validated MaxEnt models can be used to respond to a wider range of
macroecological questions regarding food webs.

Second, our analytical model can be used to generate informative priors
in Bayesian analyses of the structure of ecological networks (e.g.,
Cirtwill et al. 2019). Indeed, the probability distribution of maximum
entropy derived using MaxEnt can be used as a prior that can be
constantly updated with novel data. For instance, if we know the number
of species and the number of interactions, we can get the degree
distribution of maximum entropy, as shown in this contribution. The
degree distribution represents the probability that a species can
interact (as a predator or a prey) with a number of other species. Data
on species interactions can be used to update the prior degree
distribution to generate a more accurate posterior distribution, thus
improving our description and understanding of the system.

Third, our analytical and heuristic models can be used to make better
predictions of pairwise species interactions by constraining the space
of feasible networks, as discussed in Strydom et al. (2021). In other
words, we can use the network configuration and/or specific measures of
food-web structure derived using MaxEnt to ensure that our predictions
of interspecific interactions form feasible networks. This means that
the probability that two species interact can be conditional on the
network structure and on the probability of interactions of all other
species pairs. For that purpose, MaxEnt can be used to predict network
structure when other data is lacking.

Finally, our analytical and heuristic models can be used as alternative
null models of ecological networks to better understand and identify the
ecological processes driving food-web structure. Indeed, these
mechanisms can be better described when analyzing the deviation of
empirical data from MaxEnt predictions. A strong deviation would
indicate that ecological mechanisms not encodes in the statistical
constraints is at play for the system at hand. If deviations are
systematic, the maximum entropy model might need to be revised to
include appropriate ecological constraints. This revision process helps
us reflect on and identify what really constrains food-web structure.

The principle of maximum entropy can thus be applied for both the
prediction and understanding of natural systems. Therefore, the model
interpretation depends on how we use it. It can be used as a baseline
distribution to identify the ecological constraints organizing natural
systems. It can also be used as predictions of ecological systems. This
distinction between understanding and predicting is essential when using
and interpreting MaxEnt models.

\hypertarget{final-remarks}{%
\subsection{Final remarks}\label{final-remarks}}

One of the biggest challenges in using the principle of maximum entropy
is to identify the set of state variables that best reproduce empirical
data. We found that the number of species and the number of interactions
are important state variables for the prediction of the joint degree
distribution. Similarly, we found that the numbers of prey and predators
for each species in a food web are important state variables for the
prediction of the network configuration. However, our predictions
overestimated the symmetry of the joint degree distribution for our
analytical model and the maximum trophic level and network diameter for
our heuristic model. We should thus continue to play the ecological
detective to find these other topological constraints that would improve
the predictions of MaxEnt models and help us understand better what
really drives food-web structure.

@@

\leavevmode\hypertarget{ref-Dickinson2010CitSci}{}%
Dickinson, Janis L., Benjamin Zuckerberg, and David N. Bonter. 2010.
{``Citizen Science as an Ecological Research Tool: Challenges and
Benefits.''} In \emph{Annual Review of Ecology, Evolution, and
Systematics, Vol 41}, edited by D. J. Futuyma, H. B. Shafer, and D.
Simberloff, 41:149--72. Palo Alto: Annual Reviews.
\url{https://doi.org/10.1146/annurev-ecolsys-102209-144636}.
